---
title: "Community Assembly and Turnover"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Community Assembly and Turnover}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "svglite",
  fig.ext = "svg"
)
```

## Overview

Understanding how communities change across space requires tools that quantify
turnover at different scales. **spacc** provides three complementary approaches:

1. **Beta distance-decay**: How does compositional similarity decay with distance?

2. **Zeta diversity**: How does the number of shared species change with the
   number of sites compared?
3. **SES null models**: Are observed patterns different from random expectations?

## Data setup

```{r data}
library(spacc)
set.seed(123)

n_sites <- 60
coords <- data.frame(
  x = runif(n_sites, 0, 100),
  y = runif(n_sites, 0, 100)
)

# Spatially structured community:
# species occurrence probability decays with distance from species' "center"
n_species <- 25
species <- matrix(0, nrow = n_sites, ncol = n_species)
centers_x <- runif(n_species, 0, 100)
centers_y <- runif(n_species, 0, 100)

for (sp in seq_len(n_species)) {
  dists <- sqrt((coords$x - centers_x[sp])^2 + (coords$y - centers_y[sp])^2)
  prob <- exp(-dists / 30)
  species[, sp] <- rpois(n_sites, lambda = prob * 3)
}
colnames(species) <- paste0("sp", seq_len(n_species))

# Build accumulation curve (used later for SES null models)
sac <- spacc(species, coords, n_seeds = 10, progress = FALSE)
```

## Beta distance-decay

The `betaDecay()` function fits models to the relationship between
compositional dissimilarity and geographic distance:

```{r decay, eval = requireNamespace("ggplot2", quietly = TRUE)}
decay <- betaDecay(species, coords)
plot(decay)
```

```{r decay-summary}
summary(decay)
```

## Zeta diversity

Zeta diversity measures the number of species shared among exactly $\zeta$
sites. The rate of decline from $\zeta_1$ to higher orders indicates whether
turnover is driven by common or rare species:

```{r zeta, eval = requireNamespace("ggplot2", quietly = TRUE)}
zd <- zetaDiversity(species, coords, orders = 1:10, n_samples = 50)
plot(zd)
```

The ratio plot reveals whether turnover follows an exponential (stochastic)
or power-law (niche-driven) pattern:

```{r zeta-ratio, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(zd, type = "ratio")
```

## SES null models

Standardized effect sizes compare observed diversity patterns against null
expectations. A positive SES indicates higher diversity than expected
(overdispersion); negative SES indicates clustering.

```{r ses, eval = requireNamespace("ggplot2", quietly = TRUE)}
result <- ses(sac, species, coords,
              null_model = "curveball", n_perm = 19,
              progress = FALSE)
print(result)
```

```{r ses-plot, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(result, type = "curve")
```

The histogram shows the distribution of SES values across accumulation steps:

```{r ses-hist, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(result, type = "histogram")
```

## Combining the three approaches

These tools answer complementary questions:

| Approach | Question | Key output |
|----------|----------|------------|
| Distance-decay | How fast does similarity decline? | Decay rate, half-distance |
| Zeta diversity | Common vs. rare species turnover? | Decline ratio, power/exp fit |
| SES null models | Is the pattern non-random? | SES values, p-values |

Together they provide a comprehensive view of community assembly
mechanisms operating in the study region.

## References

- Baselga, A. (2010). Partitioning the turnover and nestedness components
  of beta diversity. *Global Ecology and Biogeography*, 19, 134-143.
- Hui, C. & McGeoch, M.A. (2014). Zeta diversity as a concept and metric
  that unifies incidence-based biodiversity patterns. *American Naturalist*,
  184, 684-694.
- Strona, G., Nappo, D., Boccacci, F., Fattorini, S. & San-Miguel-Ayanz, J.
  (2014). A fast and unbiased procedure to randomize ecological binary
  matrices with fixed row and column totals. *Nature Communications*, 5, 4114.
