---
title: "Extrapolation and Species-Area Models"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extrapolation and Species-Area Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "svglite",
  fig.ext = "svg"
)
```

## Overview

Most surveys stop before the last species is found. The observed
species count is then a lower bound, and the gap between it and the
true richness depends on how the curve was approaching its plateau when
sampling ended. spacc estimates that plateau three ways: by fitting an
asymptotic model to the mean accumulation curve, by extrapolating along
the sample-coverage axis, and by relating diversity to cumulative area.
This vignette covers the theory behind each approach, the exact model
equations as they are coded in `extrapolate()`, how to judge whether a
fit can be trusted, and when extrapolation should be avoided altogether.

The running example simulates a community with a known number of
species. The ground truth is `r 50` species spread over 100 sites, so
every estimate below can be checked against a value we actually control,
which is rarely possible with field data.

## Algorithm overview

`extrapolate()` takes a fitted `spacc` object, reduces it to the mean
accumulation curve, and fits one nonlinear model to that curve by
least squares through `stats::nls()`. The asymptotic parameter of the
model is then read off as the estimate of total richness. Six model
forms are available, selected through the `model` argument:
`"michaelis-menten"`, `"lomolino"`, `"asymptotic"`, `"weibull"`,
`"logistic"`, and `"evt"`. The first five all flatten toward a single
ceiling and differ in how sharply they bend and where the bend sits.

The **Michaelis-Menten** form (the default) rises steeply and decelerates
smoothly, reaching half its ceiling at a fixed effort. It is the
hyperbolic curve familiar from enzyme kinetics, transplanted to
accumulation. The **asymptotic exponential** approaches its ceiling at a
constant relative rate and has no inflection, so it is the most
conservative of the five when the curve is already near saturation. The
**logistic** is sigmoidal: it has an inflection point and an initial slow
phase, which suits curves that start gently before accelerating. The
**Weibull** form is a flexible generalisation that can be concave or
sigmoidal depending on its shape parameter, and it nests the exponential
as a special case. The **Lomolino** model is a sigmoidal curve on a
logarithmic effort axis, designed by Lomolino (2000) specifically for
species-area data and able to capture a long slow tail.

The **EVT** model is different in kind. Following Borda-de-Agua et al.
(2025), it represents the species-area relationship as a two-component
Weibull mixture rather than a single saturating curve. The motivation is
that empirical SARs are often triphasic: a steep small-scale phase, a
shallower intermediate phase, and a second steepening at very large
extents. A single asymptotic curve cannot bend twice, so the mixture
sums two Weibull components with separate scale and shape parameters and
a mixing weight, which lets the fitted curve change curvature more than
once. This model is fit with the bounded `"port"` algorithm and falls
back to a single Weibull when fewer than ten points are available.

The five single-curve models differ in how much flexibility they spend. A model
with more free parameters can follow the observed curve more closely, but
the extra freedom is spent describing the sampled range, not constraining
the ceiling beyond it. The asymptotic exponential, with two parameters
and no inflection, is the least flexible and so the most stable when
extrapolated, at the cost of forcing a particular approach to the plateau.
The Weibull and Lomolino forms add a third parameter that adjusts
curvature, which helps when the curve has an unusual shape but loosens the
grip on the asymptote. spacc reads the asymptote off parameter $a$ in all
six models, which keeps the interface uniform: whatever curve shape is
chosen, the estimate of total richness and its interval come from the same
slot.

Coverage-based extrapolation, exposed through `extrapolateCoverage()`,
abandons the model-fitting route entirely. Instead of asking what the
curve does as effort grows without bound, it asks what richness would be
at a target sample completeness, using the asymptotic estimators of Chao
et al. (2014). Sample coverage is the fraction of the community's total
abundance carried by the species already detected, so standardising to a
fixed coverage compares communities at equal completeness rather than
equal effort. The mechanism is rarity-driven: when many species are
singletons the coverage is low and the extrapolation deficit large,
whereas a sample dominated by species seen many times is already nearly
complete and has little room to grow. The diversity-area relationship,
fit by `dar()`, is the third route: it plots Hill numbers against
cumulative area (Ma 2018) and so describes how diversity scales spatially
across all orders, with the classical SAR recovered as the special case
of order zero.

The three routes answer slightly different questions and suit different
data. Model fitting suits a single well-sampled community where a tidy
asymptote with a confidence interval is the deliverable. Coverage
extrapolation suits comparisons across communities of unequal sampling
intensity, where the fair common ground is completeness rather than
effort. The DAR suits questions about scaling itself, where the interest
is in the exponent that links diversity to area rather than in any single
ceiling. spacc keeps all three behind a common `spacc` pipeline so the
same accumulation engine feeds each one.

## Data setup

```{r simulate}
library(spacc)

set.seed(42)
n_sites <- 100
n_species <- 50

coords <- data.frame(
  x = runif(n_sites, 0, 100),
  y = runif(n_sites, 0, 100)
)

species <- matrix(0L, n_sites, n_species)
for (sp in seq_len(n_species)) {
  cx <- runif(1, 10, 90)
  cy <- runif(1, 10, 90)
  lambda <- 4 * exp(-0.0008 * ((coords$x - cx)^2 + (coords$y - cy)^2))
  species[, sp] <- rpois(n_sites, lambda)
}
colnames(species) <- paste0("sp", seq_len(n_species))
pa <- (species > 0) * 1L
```

Each species has a spatial centre and a Poisson abundance that decays
with distance from it, so range edges thin out the way real ranges do.
The true richness is `n_species` (`r 50`), and that number anchors every
estimate that follows.

## Mathematical detail

Each model is written below with $S$ for cumulative species count, $n$
for sampling effort (number of sites accumulated), and a model-specific
parameter set. In every case the parameter $a$ is the asymptote, the
estimated total richness toward which the curve flattens.

The **Michaelis-Menten** model is

$$S = \frac{a\, n}{b + n},$$

where $a$ is the asymptote and $b$ is the effort at which $S$ reaches
$a/2$. Larger $b$ means a slower approach to the ceiling. The
**asymptotic exponential** is

$$S = a\bigl(1 - e^{-b n}\bigr),$$

with asymptote $a$ and rate constant $b > 0$ controlling how quickly the
deficit $a - S$ shrinks. The **logistic** model is

$$S = \frac{a}{1 + e^{-b (n - c)}},$$

where $a$ is the asymptote, $b$ is the steepness of the rise, and $c$ is
the effort at the inflection point where the curve is half its ceiling.
The **Weibull** model is

$$S = a\Bigl(1 - e^{-(n/b)^{c}}\Bigr),$$

with asymptote $a$, scale $b$ setting the effort axis, and shape $c$
governing curvature: $c = 1$ recovers the exponential, $c > 1$ gives a
sigmoidal rise, and $c < 1$ a sharper early climb. The **Lomolino**
model is

$$S = \frac{a}{1 + b^{\,\log(c / n)}},$$

where $a$ is the asymptote, $c$ locates the curve along the
log-transformed effort axis, and $b$ controls the slope of the
sigmoidal rise in log space. Because the effort enters as $\log(c/n)$,
the Lomolino curve stretches the small-effort region and compresses the
large-effort tail, which is why it tracks species-area data with a long
slow approach to the plateau better than the curves written on a linear
effort axis.

Starting values matter for all of these because `nls()` solves a
nonlinear least-squares problem iteratively from an initial guess.
`extrapolate()` sets the starting asymptote to $1.2$ times the observed
maximum, since the true ceiling lies above what has been seen, and sets
the half-saturation effort to the point where the observed curve first
crosses half its maximum. Reasonable starts of this kind keep the
solver inside the basin of the correct solution; poor starts are a
common reason for non-convergence on awkward curves.

The **EVT** mixture combines two Weibull components,

$$S = a\Bigl[w\bigl(1 - e^{-(n/b_1)^{c_1}}\bigr) + (1 - w)\bigl(1 - e^{-(n/b_2)^{c_2}}\bigr)\Bigr],$$

where $a$ is the shared asymptote, $w \in (0,1)$ is the mixing weight
between the two components, $b_1, b_2$ are their scales, and $c_1, c_2$
their shapes. The two components can capture two distinct phases of
curvature, which is what lets the mixture follow a triphasic SAR that a
single saturating curve flattens out. In all six models the asymptote
$a$ is the quantity of interest: it estimates the richness the community
would reveal under unlimited sampling, so the deficit $a - S_{\text{obs}}$
is the number of species the survey is predicted to have missed.

## Asymptotic extrapolation

`extrapolate()` fits an asymptotic model to the mean accumulation curve
to estimate total species richness:

```{r sac}
sac <- spacc(pa, coords, n_seeds = 30, progress = FALSE)
```

A single call returns one fitted model. The default is
Michaelis-Menten, but any of the six names can be passed. The returned
`spacc_fit` object carries the asymptote, its confidence interval, the
AIC, and the underlying `nls` fit, which is what makes `coef()`,
`confint()`, and `predict()` work on it. By default the interval is a
calibrated bootstrap across the seed curves rather than the `nls` profile
interval; the difference and the reasons for it are taken up below.

### Model comparison

spacc supports six asymptotic models. Compare them to select the best
fit:

```{r models}
models <- c("michaelis-menten", "lomolino", "asymptotic", "weibull", "logistic")
fits <- lapply(models, function(m) extrapolate(sac, model = m, interval = "none"))
names(fits) <- models

# Compare AIC
data.frame(
  model = models,
  asymptote = sapply(fits, function(f) round(f$asymptote, 1)),
  AIC = sapply(fits, function(f) round(f$aic, 1))
)
```

The asymptote estimates spread across models even when all fit the same
curve, which is the central caution of model-based extrapolation: the
ceiling is an extrapolated quantity, and models that agree closely over
the observed range can still disagree about where they level off. The
spread across these five asymptotes, set against the known true value of
`r 50`, shows directly how much model choice matters here.

```{r plot-best, fig.cap = "Best-fitting asymptotic model with bootstrap band."}
best_name <- names(fits)[which.min(sapply(fits, function(f) f$aic))]
best <- extrapolate(sac, model = best_name, progress = FALSE)
plot(best) +
  ggplot2::theme(
    panel.background = ggplot2::element_rect(fill = "transparent"),
    plot.background = ggplot2::element_rect(fill = "transparent")
  )
```

The solid line is the observed mean curve, the dashed line the fitted
model extended past the data, the shaded band the bootstrap prediction
interval, and the dotted horizontal line the estimated asymptote.

### EVT model

The Extreme Value Theory model (Borda-de-Agua et al. 2025) uses a
two-component Weibull mixture to capture the triphasic pattern
observed in empirical species-area relationships:

```{r evt}
fit_evt <- extrapolate(sac, model = "evt")
fit_evt
```

The EVT model requires sufficient data points (>10) and works best with
real ecological data exhibiting a clear triphasic SAR pattern. With only
100 sites and a single simulated phase, the mixture has little extra
structure to find, so it tends to mimic a single Weibull here. Its
advantage appears on large multi-scale datasets where the curve genuinely
bends more than once.

## Fitting and fit quality

`compareModels()` fits every model in one call and ranks them by AIC,
adding BIC, delta-AIC, Akaike weights, parameter counts, and a
convergence flag. This is the recommended entry point when the goal is
model selection rather than a single pre-chosen form.

```{r compare-models}
cm <- compareModels(sac, progress = FALSE)
cm
```

The printed table is sorted with the best model first and starred. The
data frame behind it, reachable with `as.data.frame()`, holds the same
columns for further use:

```{r compare-table}
as.data.frame(cm)[, c("model", "AIC", "delta_AIC", "AIC_weight", "converged")]
```

Models within about 2 delta-AIC units of the best are usually treated as
indistinguishable on these data; the Akaike weight turns that into a
probability that each model is the best in the set. A single model with
a weight near 1 indicates a clear winner, whereas weights spread evenly
across several models warn that the asymptote is poorly identified and
any single point estimate is fragile.

The chosen fit exposes its parameters and intervals through the standard
extractors. `coef()` returns the fitted parameters and `confint()` the
asymptote interval. By default `confint()` returns the bootstrap interval
stored on the fit; `method = "profile"` recovers the `nls`
profile-likelihood interval for comparison:

```{r coef-confint}
coef(best)
confint(best)
confint(best, method = "profile")
```

The profile interval is computed from the residuals of a single fit to
the mean curve. Because averaging many seeds makes that curve very smooth,
those residuals are tiny and the profile interval is correspondingly
narrow, ignoring the variability between seed orderings. The bootstrap
interval refits the model across resampled seed curves, so it propagates
that across-seed variability and is the wider, more honest summary of how
well the asymptote is pinned down.

`extrapolate()` also reports goodness of fit directly. The fit carries the
residual root-mean-square error over the observed range, and `print()`
shows it together with the asymptote, the nonparametric `chao2()` and
`iChao2()` estimates, and an effort range within which predictions are
reasonable:

```{r residuals}
best$gof$rmse
best
```

The fit goes through `nls()`, which can fail to converge when the curve
is short, noisy, or far from saturated; the bounded `"port"` algorithm
used for EVT is more stable but still not guaranteed. When fitting fails,
`extrapolate()` returns an object with `NA` asymptote rather than
erroring, and `compareModels()` marks that row as not converged so the
remaining models can still be ranked.

A small residual confirms that the chosen model tracks the observed range,
but it says nothing about the extrapolated part of the curve, where there
is no data to compute a residual against. Two models can have nearly
identical residuals over the sampled effort and still place their ceilings
far apart, which is why the AIC table and the asymptote interval carry more
weight than the in-sample fit statistics.

A caution applies to the asymptote itself. On clustered or strongly
under-sampled presence-absence data the parametric asymptote tends to
overestimate true richness, sometimes by tens of percent, because the
curve is far from saturation. The bootstrap interval quantifies the
uncertainty of the fitted asymptote, but it is centred on that possibly
biased estimate and is not a calibrated interval for true total richness.
`extrapolate()` warns when the asymptote runs well above the observed
richness or disagrees with `chao2()`, and for calibrated total-richness
intervals the nonparametric estimators in the richness-estimation vignette
are the better tool. Using both routes and checking that they roughly
agree is more defensible than trusting the asymptote alone.

## Tuning

Three controls shape the quality of an extrapolation. The first is the
number of points the curve provides. The EVT mixture has six free
parameters and needs more than ten accumulation points before it will
fit at all; below that threshold `extrapolate()` falls back to a single
Weibull and warns. With 100 sites there is ample room, but on a
20-plot survey the mixture should not be attempted.

The second is `n_seeds`, the number of random starting points averaged
into the mean curve. Each seed traces one accumulation order; more seeds
smooth the mean and tighten the confidence band, which steadies the fit
because `extrapolate()` works on the mean. The effect is visible in the
asymptote estimate as the seed count rises:

```{r seeds}
sapply(c(5, 20, 50), function(k) {
  s <- spacc(pa, coords, n_seeds = k, progress = FALSE)
  round(extrapolate(s, model = "michaelis-menten")$asymptote, 1)
})
```

The estimate stabilises as seeds accumulate; the gain from 20 to 50 is
typically smaller than from 5 to 20, so 30 to 50 seeds is a reasonable
default for a stable mean. The third control is the model itself, and
`compareModels()` already automates that choice by ranking the whole set
rather than committing to one form up front.

## Coverage extrapolation and DAR

`extrapolateCoverage()` estimates richness at coverage levels beyond
the observed maximum, using asymptotic estimators (Chao et al. 2014):

```{r cov-extrap}
cov <- spaccCoverage(species, coords, n_seeds = 20, progress = FALSE)

ext <- extrapolateCoverage(cov, target_coverage = c(0.95, 0.99, 1.0), q = 0)
ext
```

Targets below the observed coverage are interpolated; targets above it
invoke the Chao1-based extrapolation for `q = 0`, with separate
estimators for the Shannon (`q = 1`) and Simpson (`q = 2`) orders. The
printed output reports observed coverage and richness alongside the
extrapolated values so the size of the extrapolation step is visible.

```{r plot-cov-extrap, fig.cap = "Coverage-based extrapolation of species richness."}
plot(ext) +
  ggplot2::theme(
    panel.background = ggplot2::element_rect(fill = "transparent"),
    plot.background = ggplot2::element_rect(fill = "transparent")
  )
```

This is particularly useful when comparing communities with different
abundances. Standardizing by coverage ensures equal completeness rather
than equal sample size, so a richer but less thoroughly sampled
community is not penalised for its lower count.

`dar()` fits the relationship between diversity (Hill numbers) and cumulative
area (Ma 2018), generalising the classical species-area relationship to every
Hill number order:

```{r dar, eval = requireNamespace("sf", quietly = TRUE)}
dar_result <- dar(species, coords, q = c(0, 1, 2), n_seeds = 20, progress = FALSE)
dar_result
```

```{r plot-dar, fig.cap = "Diversity-area relationship for q = 0, 1, 2.", eval = requireNamespace("sf", quietly = TRUE)}
plot(dar_result) +
  ggplot2::theme(
    panel.background = ggplot2::element_rect(fill = "transparent"),
    plot.background = ggplot2::element_rect(fill = "transparent")
  )
```

The DAR captures how diversity scales with area. For `q = 0` it reduces
to the classical SAR; for `q = 1` it tracks how common-species diversity
scales, and for `q = 2` how dominant-species diversity scales. Curves
that diverge across orders indicate that evenness, not just richness,
changes with spatial extent. If the three curves rise in parallel,
evenness is roughly constant across scales and richness alone carries the
spatial signal; if the higher-order curves flatten earlier than the
richness curve, common and dominant species saturate first while rare
species keep accumulating, which is the usual pattern for aggregated
communities. The area axis is built from the cumulative convex hull of
the accumulated sites by default, so it requires `sf`; passing
`area_method = "count"` substitutes the site count as an area proxy when
`sf` is not available.

Reading the DAR as a power law, $D = c\,A^{z}$ with diversity $D$, area
$A$, intercept $c$, and scaling exponent $z$, gives a compact way to
compare orders: the slope $z$ on log-log axes measures how fast each
facet of diversity grows with area. A steep $z$ at order zero with a
shallow $z$ at order two is the signature of a community where rare
species drive the area effect, which is exactly the case where a richness
estimate from a small plot will most badly understate the regional total.

## Predicting richness at new effort levels

Use `predict()` on a fitted extrapolation model to read the curve at any
effort, inside or beyond the sampled range. Point predictions come back
as a plain vector:

```{r predict}
predict(best, n = c(50, 100, 200))
```

With `interval = "bootstrap"` the same call returns a prediction band
built from the bootstrap refits:

```{r predict-interval}
predict(best, n = c(50, 100, 200), interval = "bootstrap")
```

The first value falls within the observed 100 sites and so is an
interpolation, which is well supported. The prediction at 200 doubles the
effort and is a genuine extrapolation that should be read as
model-dependent. `predict()` warns once the requested effort runs past
roughly two and a half times the maximum observed effort, the common
rarefaction rule of thumb; past that range different models that fit the
data equally well diverge sharply, and the prediction band, not the point
estimate, is the honest summary.

## Extrapolating to a larger area

`extrapolate()` extends the curve in sample count: a prediction at 200
sites describes a survey twice the size of the one actually run, within
the same spatial extent. A different question is how many species a region
*larger* than the sampled extent would hold. `extrapolateArea()` answers
it with the total-species (T-S) curve of Ugland, Gray and Ellingsen
(2003). Sites are partitioned into spatial subareas; for each number of
subareas the expected total richness of random combinations is plotted
against their combined convex-hull area; an asymptotic model is fitted to
that species-area curve and extended to a target area.

```{r area}
area_fit <- extrapolateArea(pa, coords, n_subareas = 8,
                            model = "michaelis-menten", R = 100,
                            seed = 1, progress = FALSE)
area_fit
```

The observed area is the convex hull of the sampled points; the target
area defaults to twice that. The interval is a site bootstrap, recentred
on the point estimate to remove the downward bias that resampling sites
with replacement would otherwise introduce. `as.data.frame()` returns the
T-S curve, and `predict()` reads the fitted curve at any area:

```{r area-predict}
predict(area_fit, area = c(area_fit$observed_area, 2 * area_fit$observed_area),
        interval = "bootstrap")
```

As with sample-count extrapolation, the fitted ceiling is model-dependent
and most trustworthy for areas not far beyond the one sampled.

## Comparison to alternatives

The coverage route in spacc shares its statistical machinery with
iNEXT, which popularised coverage-based rarefaction and extrapolation
through Hill numbers (Chao et al. 2014). Both standardise communities by
sample completeness rather than raw effort, and both lean on the same
Chao-type asymptotic estimators when extrapolating past the observed
data. The difference is structural. iNEXT treats samples as exchangeable
and accumulates them without geography, whereas spacc accumulates sites
in spatial order through its nearest-neighbour engine, so the coverage
curve reflects how completeness builds up across space rather than across
an arbitrary sample order. For spatially aggregated plot data the
sample-based Chiu (2023) estimator, available through the `coverage`
argument of `spaccCoverage()`, is the more appropriate coverage estimator
than the individual-based default.

The model-based route in `extrapolate()` is a complement, not a rival, to
the coverage route. Its strength is interpretability: the asymptote is a
single number with a confidence interval, and the AIC table makes model
uncertainty explicit. Its limit is that the estimate is only as good as
the model shape, and an unsaturated curve underdetermines the ceiling, so
the asymptotes can scatter widely even when every model fits the observed
range. Coverage extrapolation sidesteps the shape assumption but pays for
it with reliance on singleton and doubleton counts, which are themselves
noisy in small samples. Using both and checking that they roughly agree
is more defensible than trusting either alone.

A second contrast is in what the two routes extrapolate against. The
fitted models extrapolate against effort, so a prediction at 500 sites is
a statement about a survey five times the size of the one actually run.
Coverage extrapolation extrapolates against completeness, so a target of
99% coverage is a statement about a survey complete enough to leave only
1% of abundance undetected, regardless of how many sites that takes. The
coverage target is bounded above by 1, which makes the extrapolation step
finite and self-limiting, whereas the effort axis has no natural ceiling
and invites predictions far past anything the data can support. For that
reason the coverage route degrades more gracefully when pushed, while the
model route gives the more transparent single estimate when the curve is
near saturation. The two also differ in their data demands: model fitting
needs the full accumulation curve, while coverage estimators need only the
final abundance vector and its rare-species counts.

## Practical guidance

A few rules of thumb keep extrapolation honest. First, fit the whole
model set with `compareModels()` rather than committing to one form;
treat models within 2 delta-AIC units as tied and report the range of
their asymptotes, not a single value. Second, do not extrapolate past
about twice the maximum observed effort, where models that fit the data
equally well begin to diverge. Third, require at least 10 accumulation
points before fitting any model and reserve the six-parameter EVT mixture
for datasets with more than about 30 points and a visibly multi-phase
curve.

The table below summarises which model to reach for first.

| Situation | Suggested model | Why |
|---|---|---|
| Curve near saturation, simple shape | asymptotic exponential | fewest parameters, conservative ceiling |
| Smooth decelerating rise, no clear inflection | Michaelis-Menten | robust default, two parameters |
| Sigmoidal rise with a slow start | logistic | inflection point captures the lag |
| Flexible single-phase SAR | Weibull | shape parameter adapts curvature |
| Classic species-area data, long tail | Lomolino | log-axis sigmoid built for SARs |
| Large multi-scale data, triphasic curve | EVT mixture | two components bend the curve twice |

Minimum data is the gate before any of this applies. Below 10
accumulation points no model should be fit; between 10 and 30 points the
two- and three-parameter single curves are usable but the EVT mixture is
not, and the asymptote interval should be reported in full rather than
collapsed to a point. Above 30 points with a visibly flattening curve all
six models become candidates and `compareModels()` can do the selecting.
A useful diagnostic is the ratio of observed richness to the fitted
asymptote, printed by `print()` on a fit: when the observed count is
already 90% or more of the estimate the extrapolation is a modest
correction, but when it is below about 70% the survey is far from
complete and the ceiling rests heavily on the model.

There are cases where the answer is to not extrapolate. When the
observed curve is still rising steeply at the final site, the asymptote
is essentially unconstrained, the asymptotes from different models will
diverge by large margins, and any point estimate is a guess dressed as a
number. In that situation the saturation summary from `summary()` and the
slope at the curve's end are more honest than a fitted ceiling, and the
correct conclusion is that more sampling is needed before total richness
can be estimated at all. Heavily aggregated communities with many
single-site species are a second warning case, because the rare-species
counts that drive both the model deficit and the coverage estimators are
then dominated by sampling noise.

## References

- Borda-de-Agua, L., Whittaker, R.J., Cardoso, P., et al. (2025).
  Extreme value theory explains the topography and scaling of the
  species-area relationship. Nature Communications, 16, 5346.
- Chao, A. & Jost, L. (2012). Coverage-based rarefaction and
  extrapolation: standardizing samples by completeness rather than size.
  Ecology, 93, 2533-2547.
- Chao, A., Gotelli, N.J., Hsieh, T.C., et al. (2014). Rarefaction and
  extrapolation with Hill numbers: a framework for sampling and
  estimation in species diversity studies. Ecological Monographs, 84,
  45-67.
- Chiu, C.-H. (2023). A sample-based estimator for sample coverage.
  Ecology, 104, e4099.
- Lomolino, M.V. (2000). Ecology's most general, yet protean pattern:
  the species-area relationship. Journal of Biogeography, 27, 17-26.
- Ma, Z. (2018). Generalizing the species-area relationship with
  diversity-area relationships. Ecology and Evolution, 8, 8645-8655.
```
