---
title: "Getting Started with spacc"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spacc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "svglite",
  fig.ext = "svg"
)
```

## Overview

**spacc** computes spatial species accumulation curves (SACs) using
nearest-neighbour algorithms with a C++ backend for speed. Unlike
classical random SACs, spatial SACs respect the geographic arrangement of
sampling sites, revealing how species richness accumulates as you expand
outward from a focal point.

## Simulating Example Data

```{r simulate}
library(spacc)

set.seed(42)
n_sites <- 80
n_species <- 40

coords <- data.frame(
  x = runif(n_sites, 0, 100),
  y = runif(n_sites, 0, 100)
)

# Simulate presence/absence with spatial structure
species <- matrix(0L, n_sites, n_species)
for (sp in seq_len(n_species)) {
  cx <- runif(1, 20, 80)
  cy <- runif(1, 20, 80)
  prob <- exp(-0.001 * ((coords$x - cx)^2 + (coords$y - cy)^2))
  species[, sp] <- rbinom(n_sites, 1, prob)
}
colnames(species) <- paste0("sp", seq_len(n_species))
```

## Basic Spatial Accumulation Curve

The core function `spacc()` computes kNN-based accumulation curves from
multiple random starting sites (seeds):
```{r basic-sac}
sac <- spacc(species, coords, n_seeds = 30, method = "knn", progress = FALSE)
sac
```

```{r plot-sac, fig.cap = "Spatial species accumulation curve with 95% confidence ribbon."}
plot(sac)
```

The grey ribbon shows variability across starting points — wider ribbons
indicate that species richness depends on *where* you start sampling.

## Comparing Methods

spacc supports several accumulation methods:

- **knn**: k-Nearest Neighbour (default, fast)

- **kncn**: k-Nearest Centroid Neighbour (geometrically centred expansion)

- **random**: Classical random accumulation (no spatial structure)

```{r compare-methods}
sac_kncn <- spacc(species, coords, n_seeds = 30, method = "kncn", progress = FALSE)
sac_rand <- spacc(species, coords, n_seeds = 30, method = "random", progress = FALSE)
```

Combine curves for comparison:
```{r combine, fig.cap = "Comparison of accumulation methods."}
combined <- c(knn = sac, kncn = sac_kncn, random = sac_rand)
plot(combined)
```

## Extrapolation

Fit an asymptotic model to estimate total species richness:
```{r extrapolate}
fit <- extrapolate(sac, model = "lomolino")
fit
```

```{r plot-fit, fig.cap = "Fitted Lomolino curve with asymptote estimate."}
plot(fit)
```

## Pre-computing Distances

For repeated analyses on the same sites, pre-compute the distance matrix once:
```{r distances}
d <- distances(coords, method = "euclidean")

sac1 <- spacc(species[, 1:20], d, n_seeds = 30, progress = FALSE)
sac2 <- spacc(species[, 21:40], d, n_seeds = 30, progress = FALSE)
```

## Comparing Curves

Test whether two accumulation curves differ significantly:
```{r compare-test}
comp <- compare(sac1, sac2, method = "permutation", n_perm = 199)
comp
```

## Next Steps
- [Diversity](diversity.html): Hill numbers, beta diversity, phylogenetic
  and functional diversity accumulation
- [Extrapolation](extrapolation.html): Coverage-based extrapolation, EVT
  models, diversity-area relationships
- [Spatial Analysis](spatial-analysis.html): Endemism curves, fragmentation
  analysis, sampling-effort correction
