---
title: "Rarefaction and Standardization"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rarefaction and Standardization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "svglite",
  fig.ext = "svg"
)
```

## Overview

Comparing diversity across sites with unequal sampling effort requires
standardization. **spacc** offers three approaches:

1. **Individual-based rarefaction**: Subsample to equal number of individuals

2. **Coverage-based rarefaction**: Standardize by sample completeness

3. **Spatial coverage tracking**: Monitor coverage along accumulation curves

This vignette compares these methods and provides guidance on when to use each.

## Data setup

```{r data}
library(spacc)
set.seed(42)

n_sites <- 60
coords <- data.frame(x = runif(n_sites), y = runif(n_sites))

# Variable total abundances across sites (realistic uneven sampling)
lambdas <- rep(c(1, 3, 5), each = 20)
species <- matrix(0, nrow = n_sites, ncol = 20)
for (i in seq_len(n_sites)) {
  species[i, ] <- rpois(20, lambda = lambdas[i])
}
colnames(species) <- paste0("sp", 1:20)
```

## Individual-based rarefaction

The `rarefy()` function subsamples curves to a common number of
individuals, removing the abundance bias:

```{r rarefy, eval = requireNamespace("ggplot2", quietly = TRUE)}
# Rarefy to minimum observed abundance
rare <- rarefy(species)
print(rare)
```

```{r rarefy-plot, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(rare)
```

You can also rarefy Hill number curves:

```{r rarefy-hill, eval = requireNamespace("ggplot2", quietly = TRUE)}
# Rarefy at different diversity orders
rare_q1 <- rarefy(species, q = 1)
rare_q2 <- rarefy(species, q = 2)
print(rare_q1)
```

## Coverage-based standardization

Coverage-based rarefaction (Chao & Jost 2012) standardizes by sample
completeness rather than sample size. This is often preferred because
equal coverage means equal proportional representation of the community:

```{r coverage, eval = requireNamespace("ggplot2", quietly = TRUE)}
cov_result <- spaccCoverage(species, coords, n_seeds = 10, progress = FALSE)
plot(cov_result)
```

### Interpolation at fixed coverage

```{r interpolate}
interp <- interpolateCoverage(cov_result, target = c(0.8, 0.9, 0.95))
print(interp)
```

### Extrapolation beyond observed

```{r extrapolate}
extrap <- extrapolateCoverage(cov_result, target_coverage = 0.99, q = 0)
print(extrap)
```

## Combined Hill + Coverage analysis

The `spaccHillCoverage()` function tracks both Hill numbers and coverage
simultaneously, enabling standardization across q orders:

```{r hill-coverage, eval = requireNamespace("ggplot2", quietly = TRUE)}
hc <- spaccHillCoverage(species, coords, q = c(0, 1, 2),
                         target_coverage = 0.9,
                         n_seeds = 10, progress = FALSE)
print(hc)
```

```{r hill-coverage-plot, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot(hc, xaxis = "coverage")
```

## When to use which method

| Method | Best for | Limitation |
|--------|----------|------------|
| Individual-based | Simple comparisons | Sensitive to abundance distribution |
| Coverage-based | Uneven sampling | Requires abundance data |
| Hill + Coverage | Multi-order standardization | Computationally heavier |

**Rules of thumb:**

- Use individual-based rarefaction when total abundances are the primary
  source of variation and you want a simple, well-understood correction.
- Use coverage-based methods when sites differ in sampling completeness
  and you want to compare at equal representativeness.
- Use Hill + Coverage when you need standardized comparisons across
  multiple diversity orders (q = 0, 1, 2) simultaneously.

## References

- Chao, A. & Jost, L. (2012). Coverage-based rarefaction and extrapolation:
  standardizing samples by completeness rather than size. *Ecology*, 93,
  2533-2547.
- Chao, A., Gotelli, N.J., Hsieh, T.C., Sander, E.L., Ma, K.H., Colwell,
  R.K. & Ellison, A.M. (2014). Rarefaction and extrapolation with Hill
  numbers. *Ecological Monographs*, 84, 45-67.
