---
title: "Richness Estimation and Completeness"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Richness Estimation and Completeness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "svglite",
  fig.ext = "svg"
)
```

## Overview

How many species actually exist in a region? Observed richness always underestimates
true richness because rare species go undetected. **spacc** provides a family of
non-parametric richness estimators that use the frequency of rare species to infer
the number of unseen ones.

This vignette covers:

1. Classical estimators (Chao1/2, ACE, jackknife, bootstrap)
2. Improved estimators (iChao1, iChao2)
3. Sample completeness profiles

## Data setup

```{r data}
library(spacc)
set.seed(42)

# 50 sites, 30 species, moderate rarity
species <- matrix(rpois(50 * 30, lambda = 2), nrow = 50)
colnames(species) <- paste0("sp", 1:30)
```

## Non-parametric richness estimators

All estimators return a `spacc_estimate` object with point estimate, standard
error, and 95% confidence interval.

### Abundance-based: Chao1 and ACE

```{r chao1}
chao1(species)
ace(species)
```

### Incidence-based: Chao2 and jackknife

```{r chao2}
chao2(species)
jackknife(species, order = 1)
jackknife(species, order = 2)
```

### Bootstrap estimator

```{r bootstrap}
bootstrap_richness(species, n_boot = 100)
```

### Comparison table

```{r comparison}
estimators <- list(
  chao1(species),
  chao2(species),
  ace(species),
  jackknife(species, order = 1),
  jackknife(species, order = 2),
  bootstrap_richness(species, n_boot = 100)
)

do.call(rbind, lapply(estimators, as.data.frame))
```

## Improved Chao estimators (iChao1/iChao2)

The improved estimators (Chiu et al. 2014) use tripletons (f3) and
quadrupletons (f4) to reduce negative bias, particularly for
under-sampled communities.

```{r ichao}
r_ichao1 <- iChao1(species)
r_ichao2 <- iChao2(species)

r_ichao1
r_ichao2
```

Compare with standard Chao:

```{r ichao-compare}
rbind(
  as.data.frame(chao1(species)),
  as.data.frame(r_ichao1),
  as.data.frame(chao2(species)),
  as.data.frame(r_ichao2)
)
```

The iChao estimators are always >= their standard counterparts.
When f4 or Q4 = 0, they collapse to the standard Chao1/Chao2.

## Sample completeness profile

The `completenessProfile()` function shows how sample completeness (the
proportion of the community represented) changes with accumulation:

```{r completeness, eval = requireNamespace("ggplot2", quietly = TRUE)}
coords <- data.frame(x = runif(50), y = runif(50))

cp <- completenessProfile(species, coords, n_seeds = 10, progress = FALSE)
plot(cp)
```

## References

- Chao, A. (1984). Nonparametric estimation of the number of classes in a
  population. *Scandinavian Journal of Statistics*, 11, 265-270.
- Chao, A. (1987). Estimating the population size for capture-recapture data
  with unequal catchability. *Biometrics*, 43, 783-791.
- Chiu, C.H., Wang, Y.T., Walther, B.A. & Chao, A. (2014). An improved
  nonparametric lower bound of species richness via a modified Good-Turing
  frequency formula. *Biometrics*, 70, 671-682.
- Chao, A. & Lee, S.M. (1992). Estimating the number of classes via sample
  coverage. *Journal of the American Statistical Association*, 87, 210-217.
