---
title: "Methods and Algorithms"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methods and Algorithms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "svglite",
  fig.ext = "svg"
)
library(spacc)
set.seed(1)
```

## Overview

This vignette describes what `spacc()` computes and how. A classical species
accumulation curve draws sampling units in random order; `spacc` draws them in
spatial order, expanding outward from a focal site so the curve reflects how
richness grows with surveyed *area* rather than with effort alone. The expansion
rule is a choice, and the package offers seven of them. Each rule defines an
ordering of the sites; the curve at step $k$ is the number of distinct species
seen in the first $k$ sites of that ordering. Repeating the expansion from many
starting sites turns the spread of the curves into a confidence band.

The sections below define the vocabulary, state the accumulation problem
formally, derive each expansion method from its ordering rule, give the distance
metrics those rules depend on, map the mathematics onto the function arguments,
and explain the two-tier nearest-neighbour backend that keeps the computation
fast on large datasets. The applied vignettes build on this foundation: see
`vignette("quickstart")` for the basic workflow, `vignette("diversity")` for the
Hill, beta, and coverage extensions, and `vignette("extrapolation")` for fitting
asymptotic models to the curves.

### Contents

1. [Terminology](#terminology)

2. [The spatial walk, concretely](#the-spatial-walk-concretely)

3. [Problem formulation](#problem-formulation)

4. [Expansion methods](#expansion-methods)

5. [Distance metrics](#distance-metrics)

6. [From theory to implementation](#from-theory-to-implementation)

7. [The two-tier backend](#the-two-tier-backend)

8. [Seeds and uncertainty](#seeds-and-uncertainty)

9. [Relationship to random-order curves](#relationship-to-random-order-curves)

10. [From curves to other diversity measures](#from-curves-to-other-diversity-measures)

11. [Design notes](#design-notes)

12. [References](#references)

---

## Terminology

These terms recur throughout the package documentation.

**Site.** One sampling unit: a plot, quadrat, grid cell, or locality. Sites carry
two coordinates and a row of species records.

**Occurrence matrix.** The input `x`: an $n \times m$ matrix of $n$ sites by $m$
species. Entries are presence-absence ($0/1$) or abundance (counts). The
accumulation algorithms reduce abundance to presence-absence internally, since a
curve counts whether a species is *new*, not how many individuals it has.

**Coordinates.** The input `coords`: an $n \times 2$ table of positions, supplied
as a data frame with `x` and `y` columns, an `sf` point layer, or a precomputed
`spacc_dist` object.

**Ordering.** A permutation $\pi = (\pi_1, \dots, \pi_n)$ of the site indices that
fixes the sequence in which sites enter the curve. Every expansion method is a
rule for producing an ordering.

**Accumulation curve.** The vector $S(1), \dots, S(n)$ where $S(k)$ is the number
of distinct species recorded across the first $k$ sites of an ordering. The curve
is non-decreasing and saturates at the total species count.

**Seed.** The starting site $\pi_1$. Spatial methods grow an ordering outward from
the seed, so the seed determines the curve.

**Neighbourhood.** The set of unvisited sites considered for the next step. For
nearest-neighbour expansion it is all unvisited sites; for the cone method it is
restricted to a directional wedge.

**Seed band.** The set of curves obtained by repeating an expansion from many
seeds. Its pointwise quantiles form the confidence band.

**Backend.** The data structure used to answer nearest-neighbour queries: a
precomputed distance matrix (exact) or a spatial index (k-d tree or ball tree).

---

## The spatial walk, concretely

Start with twenty sites and watch one rule build an ordering. The k-nearest-neighbour
walk begins at a seed, steps to the closest unvisited site, then to the closest
site to *that*, and so on. The path threads through the point cloud.

```{r toy-points}
pts <- data.frame(
  x = c(1, 2, 1.5, 3, 3.5, 5, 5.5, 6, 2, 4,
        7, 7.5, 8, 1, 6.5, 4.5, 2.5, 8.5, 5, 3),
  y = c(1, 1.2, 2, 1, 2.5, 1, 2, 3, 3.5, 4,
        1.5, 3, 2, 4.5, 4, 3.5, 4.8, 4, 5, 5.5)
)
```

The walk is easy to trace by hand. From the seed, repeatedly pick the nearest
site that has not yet been visited.

```{r toy-knn-order}
knn_order <- function(coords, seed) {
  n <- nrow(coords); visited <- logical(n); ord <- integer(n)
  cur <- seed; visited[cur] <- TRUE; ord[1] <- cur
  for (k in 2:n) {
    d <- sqrt((coords$x - coords$x[cur])^2 + (coords$y - coords$y[cur])^2)
    d[visited] <- Inf
    cur <- which.min(d); visited[cur] <- TRUE; ord[k] <- cur
  }
  ord
}
ord <- knn_order(pts, seed = 1)
```

Drawing the path shows how the ordering hugs the local structure: the first steps
stay inside one cluster before the walk is forced to jump to the next.

```{r toy-knn-plot, fig.width=6, fig.height=5, fig.alt="Twenty points in the plane with a path connecting them in nearest-neighbour visiting order, starting from the lower-left seed. The path stays within local clusters before jumping to reach distant points."}
plot(pts$x, pts$y, pch = 19, col = "grey70", cex = 1.4,
     xlab = "x", ylab = "y", main = "kNN walk from seed (filled = seed)")
lines(pts$x[ord], pts$y[ord], col = "#2E7D32", lwd = 2)
points(pts$x[1], pts$y[1], pch = 19, col = "#C62828", cex = 2)
text(pts$x, pts$y, labels = match(seq_len(nrow(pts)), ord), pos = 3, cex = 0.7)
```

A different seed gives a different path and a different curve. That dependence on
the starting point is not noise to be removed; it is the spatial signal. When
species are aggregated, a walk that starts inside a rich cluster climbs steeply,
a walk that starts in a sparse corner climbs slowly, and the gap between those
trajectories measures how much composition turns over across the map. The package
runs the walk from many seeds precisely to capture that spread.

---

## Problem formulation

**Input.** An occurrence matrix $X \in \{0,1,2,\dots\}^{n \times m}$ for $n$ sites
and $m$ species, and coordinates $c_1, \dots, c_n \in \mathbb{R}^2$. Let
$\mathrm{sp}(i) = \{\, j : X_{ij} > 0 \,\}$ be the set of species present at site
$i$.

**Ordering.** A method produces a permutation $\pi$ of $\{1, \dots, n\}$. For the
stochastic methods $\pi$ is a random variable whose distribution depends on the
seed and the rule.

**Accumulation function.** Given an ordering $\pi$, the curve is

$$
S_\pi(k) \;=\; \left| \, \bigcup_{t=1}^{k} \mathrm{sp}(\pi_t) \, \right|,
\qquad k = 1, \dots, n .
$$

$S_\pi$ is a step function: it rises by the count of species at $\pi_k$ that were
absent from the first $k-1$ sites, and never falls. It reaches the total richness
$S_\pi(n) = |\bigcup_i \mathrm{sp}(i)|$ regardless of the ordering, so methods
differ only in the *shape* of the approach, not the endpoint.

**Seed band.** Run the method from seeds $s_1, \dots, s_B$ (sampled with
replacement from the eligible sites), giving curves $S^{(1)}, \dots, S^{(B)}$
stacked as the $B \times n$ matrix returned in `$curves`. The summary at step $k$
is the across-seed mean and the empirical quantiles

$$
\bar S(k) = \frac{1}{B} \sum_{b=1}^{B} S^{(b)}(k), \qquad
\hat q_\alpha(k) = \text{quantile}_\alpha\!\left( S^{(1)}(k), \dots, S^{(B)}(k) \right).
$$

The $2.5\%$ and $97.5\%$ quantiles give the default band. The interval is not a
parametric formula; it is the sampling distribution of richness-at-effort induced
by varying the starting point.

---

## Expansion methods

Each method is a rule for choosing $\pi_{k+1}$ given the sites already visited.
The seven rules fall into three families: walks that chain through the cloud
(`knn`, `kncn`, `gaussian`), orderings by distance from a fixed seed (`radius`,
`cone`), and geography-free baselines (`random`, `collector`).

### k-nearest neighbour (`knn`)

From the current site $\pi_k$, move to the closest unvisited site:

$$
\pi_{k+1} \;=\; \arg\min_{j \,\notin\, \{\pi_1,\dots,\pi_k\}} d(\pi_k, j).
$$

The reference point is the *current* site, so the ordering is a connected walk
that follows local density. This is the default and the curve the other methods
are usually compared against.

### k-nearest centroid neighbour (`kncn`)

Track the centroid of the visited set and move to the unvisited site closest to
it:

$$
\bar c_k = \frac{1}{k} \sum_{t=1}^{k} c_{\pi_t}, \qquad
\pi_{k+1} = \arg\min_{j \,\notin\, \{\pi_1,\dots,\pi_k\}} \| c_j - \bar c_k \| .
$$

Because the reference point is the centroid rather than the last site, the visited
set grows as a compact blob instead of a thread. `kncn` resists the long jumps a
`knn` walk makes when it exhausts a cluster, so its early curve is smoother.

### Expanding radius (`radius`)

Sort every site by its distance from the seed and accumulate in that order:

$$
\pi \;=\; \text{argsort}_j \; d(s, j),
\qquad s = \text{seed}.
$$

The reference point is fixed at the seed for the whole curve, so the ordering
sweeps out a growing disc centred on the seed. This is the cleanest "survey
spreading outward from a point" interpretation. The related `spaccWavefront()`
function parameterises the same idea by radius instead of by site count, reporting
richness as a function of the disc radius rather than the number of sites
included.

### Gaussian-weighted walk (`gaussian`)

A soft version of `knn`. From the current site, draw the next site at random with
probability proportional to a Gaussian kernel of distance:

$$
\Pr(\pi_{k+1} = j) \;\propto\; \exp\!\left( -\frac{d(\pi_k, j)^2}{2\sigma^2} \right),
\qquad j \notin \{\pi_1,\dots,\pi_k\}.
$$

Nearby sites are favoured but not guaranteed, so the walk explores a neighbourhood
rather than always taking the single closest site. The bandwidth $\sigma$ sets how
sharp the preference is: small $\sigma$ approaches `knn`, large $\sigma$ approaches
`random`. By default $\sigma$ is the median of the non-zero pairwise distances.

### Directional cone (`cone`)

Pick a random direction $\theta$ for the seed. Sites whose bearing from the seed
falls within the half-width `cone_width` of $\theta$ are accumulated first, in
order of distance; the remaining sites follow afterwards, also by distance:

$$
\text{in-cone}(j) \iff \bigl| \angle(c_j - c_s) - \theta \bigr| \le \texttt{cone\_width}.
$$

The cone models a survey that advances along a bearing, such as a transect up a
valley or along a coastline. The default half-width is $\pi/4$ (a 90-degree wedge).

### Random order (`random`)

A uniform random permutation of the sites, independent of geography. This is the
classical accumulation curve and the null model the spatial methods are measured
against.

### Collector (`collector`)

The sites in the order they appear in the data, with no randomisation. It produces
a single curve and reproduces the "collector's curve" of the order in which a
survey was actually conducted.

### Comparing the orderings

The same points, ordered by four rules, make the differences visible. Colour
encodes visiting order from first (dark) to last (light).

```{r method-schematic, fig.width=7, fig.height=6, fig.alt="Four panels showing the same twenty points coloured by visiting order under kNN, expanding radius, random, and collector rules. The kNN and radius panels show smooth spatial gradients of colour; the random panel shows no spatial pattern; the collector panel follows data order."}
pal <- function(o) grDevices::hcl.colors(length(o), "Greens", rev = TRUE)[order(o)]
radius_order <- function(coords, seed)
  order(sqrt((coords$x - coords$x[seed])^2 + (coords$y - coords$y[seed])^2))

orders <- list(
  kNN       = knn_order(pts, 1),
  radius    = radius_order(pts, 1),
  random    = sample(nrow(pts)),
  collector = seq_len(nrow(pts))
)
op <- par(mfrow = c(2, 2), mar = c(2, 2, 2, 1))
for (nm in names(orders)) {
  rank_k <- match(seq_len(nrow(pts)), orders[[nm]])
  plot(pts$x, pts$y, pch = 19, cex = 1.6, col = pal(rank_k),
       xlab = "", ylab = "", main = nm, axes = FALSE); box()
}
par(op)
```

The `knn` and `radius` panels show a clear gradient: nearby sites are visited at
similar times. The `random` panel has no spatial pattern, and `collector` follows
whatever order the rows happen to be in.

---

## Distance metrics

The walks depend on a distance $d(i, j)$ between sites, set by the `distance`
argument.

**Euclidean.** For projected or arbitrary planar coordinates,

$$
d(i,j) = \sqrt{(x_i - x_j)^2 + (y_i - y_j)^2}.
$$

**Haversine.** For geographic coordinates given as longitude (`x`) and latitude
(`y`) in degrees, the great-circle distance in kilometres,

$$
d(i,j) = 2R \,\arcsin\!\sqrt{ \sin^2\!\tfrac{\Delta\varphi}{2}
  + \cos\varphi_i \cos\varphi_j \, \sin^2\!\tfrac{\Delta\lambda}{2} },
$$

with $\varphi$ latitude, $\lambda$ longitude, and $R = 6371$ km. Haversine needs no
projection step, so longitude-latitude data can be used directly.

**Spatiotemporal.** When sites are sampled across time, supply a `time` vector and
the distance becomes a weighted sum of a spatial and a temporal term:

$$
d(i,j) = w_{\text{space}} \, d_{\text{space}}(i,j) + w_{\text{time}} \, |t_i - t_j|.
$$

The weights `w_space` and `w_time` trade geographic against temporal proximity. A
composite distance is not a metric a spatial tree can index, so this mode always
uses the exact backend and is available for the methods that accept a distance
matrix (`knn`, `radius`, `gaussian`).

---

## From theory to implementation

The mathematical objects map onto arguments of `spacc()` directly.

| Concept | Argument | Effect |
|---|---|---|
| Ordering rule | `method` | Selects one of the seven expansion rules |
| Distance $d$ | `distance` | `"euclidean"` or `"haversine"` |
| Number of seeds $B$ | `n_seeds` | Curves in the seed band |
| Bandwidth $\sigma$ | `sigma` | Gaussian kernel width (default: median distance) |
| Cone half-width | `cone_width` | Angular wedge for `cone` (default $\pi/4$) |
| Temporal axis | `time`, `w_space`, `w_time` | Switches to composite distance |
| Fixed ordering | `order` | Supplies $\pi$ directly, bypassing `method` |
| Species split | `groups` | One curve per group, same site ordering |
| Spatial support | `support`, `include_halo` | Seeds drawn from core sites only |
| Query backend | `backend` | `"auto"`, `"exact"`, or `"kdtree"` |

A typical call states the rule, the metric, and the number of seeds:

```{r impl-call}
coords  <- data.frame(x = runif(120), y = runif(120))
species <- matrix(rbinom(120 * 40, 1, 0.25), nrow = 120)

sac <- spacc(species, coords, method = "knn", n_seeds = 100,
             progress = FALSE, seed = 42)
sac
```

Supplying `order` brings an externally computed ordering, for example one from a
sampling design or a different package, into the same band-and-plot machinery.
Each ordering is treated like a seed, so a matrix of orderings yields a band.

```{r impl-order}
ord_mat <- t(replicate(20, sample(nrow(species))))
sac_user <- spacc(species, coords, order = ord_mat, progress = FALSE)
sac_user$n_seeds
```

---

## The two-tier backend

The cost of an accumulation walk is dominated by nearest-neighbour queries. Two
backends answer them, and `backend = "auto"` chooses between them by site count.

**Exact.** Precompute the full $n \times n$ distance matrix once, then answer each
query by scanning a row. The matrix costs $O(n^2)$ memory and the scan costs
$O(n)$ per step, so a full walk is $O(n^2)$. For up to a few hundred sites this is
the fastest option and the matrix fits comfortably in memory.

**Spatial tree.** Build a spatial index and query it in roughly $O(\log n)$,
avoiding the quadratic matrix entirely. For Euclidean distances `spacc` uses a
k-d tree (via nanoflann); for haversine distances it uses a ball tree, whose
spherical-cap geometry suits great-circle distance. The build is $O(n \log n)$ and
the walk is about $O(n \log n)$, which wins decisively once $n$ grows past a few
hundred.

| Backend | Distance | Memory | Per query | Selected when |
|---|---|---|---|---|
| Exact | any | $O(n^2)$ | $O(n)$ | $n \le 500$ (auto) |
| k-d tree | Euclidean | $O(n)$ | $\approx O(\log n)$ | $n > 500$ (auto) |
| Ball tree | Haversine | $O(n)$ | $\approx O(\log n)$ | $n > 500$, haversine |

The `auto` rule switches to a tree above 500 sites. The two backends compute the
same ordering rule, so they agree on the curve up to ties; only the speed differs.

```{r backend-equiv}
e <- spacc(species, coords, method = "knn", backend = "exact",
           n_seeds = 30, progress = FALSE, seed = 7)
k <- spacc(species, coords, method = "knn", backend = "kdtree",
           n_seeds = 30, progress = FALSE, seed = 7)
c(exact_mean_end = mean(e$curves[, ncol(e$curves)]),
  kdtree_mean_end = mean(k$curves[, ncol(k$curves)]))
```

The `radius`, `gaussian`, and `cone` methods always use coordinates or a distance
matrix directly, and the spatiotemporal composite forces the exact backend because
a summed space-time distance has no tree to index.

---

## Seeds and uncertainty

A single spatial curve is one realisation. The uncertainty comes from the choice
of starting point, so `spacc` repeats the expansion from `n_seeds` seeds and reads
the spread off the resulting band. Each seed is an independent walk with no shared
state, which makes the computation embarrassingly parallel: the seeds are
distributed across threads by RcppParallel, and the per-step quantiles are taken
after all walks finish.

```{r seed-band, fig.width=7, fig.height=4.5, fig.alt="A spatial accumulation curve with a shaded confidence band. The mean curve rises and saturates; the band is widest in the early, low-effort region where the choice of starting site matters most."}
plot(sac)
```

The band is widest early and narrows as the curves converge on the shared
endpoint. Early on, the seed dominates: a walk that starts in a rich patch and one
that starts in a poor patch disagree most after a handful of sites. By the time
most of the map is covered, every walk has seen nearly every species, so the
curves meet. More seeds tighten the *estimate* of the band but do not change its
shape; the band reflects real across-start variability, not Monte Carlo error that
vanishes with more replicates.

```{r seed-count}
b50  <- spacc(species, coords, n_seeds = 50,  progress = FALSE, seed = 1)
b300 <- spacc(species, coords, n_seeds = 300, progress = FALSE, seed = 1)
mid <- round(ncol(b50$curves) / 4)
c(seeds_50  = diff(quantile(b50$curves[,  mid], c(.025, .975))),
  seeds_300 = diff(quantile(b300$curves[, mid], c(.025, .975))))
```

---

## Relationship to random-order curves

The classical accumulation curve in `vegan::specaccum()` draws sites in random or
collector order and averages over permutations. `spacc` reproduces that curve with
`method = "random"` or `method = "collector"` and adds the spatial methods on top,
so the two can be read against each other. The `compare()` function tests the gap
between any two curves by permutation, bootstrap, or area-under-curve.

```{r compare-random, fig.width=7, fig.height=4.5, fig.alt="Two accumulation curves on the same axes: the random-order curve rises faster and the spatial kNN curve lags behind it, with the shaded gap between them representing spatial turnover."}
sp <- spacc(species, coords, method = "knn",    n_seeds = 100, progress = FALSE, seed = 3)
rd <- spacc(species, coords, method = "random", n_seeds = 100, progress = FALSE, seed = 3)
cmp <- compare(sp, rd)
plot(cmp)
```

When species are aggregated, the spatial curve lags the random curve: neighbouring
sites share species, so spreading outward discovers them more slowly than mixing
distant communities at every step. The size of the gap is a measure of spatial
turnover. A spatial curve that tracks the random one signals a community with
little spatial structure; a curve that lags far behind signals strong distance
decay in composition. An existing `vegan::specaccum()` result can be imported with
`as_spacc()` to enter the same plotting and comparison machinery.

---

## From curves to other diversity measures

The expansion rule fixes an ordering; what is counted at each step is free to
change. The downstream functions reuse the spatial walk and replace the quantity
accumulated, so the spatial logic in this vignette carries through unchanged.

- `spaccHill()` accumulates Hill numbers of order $q = 0, 1, 2$ (richness, the
  exponential of Shannon entropy, the inverse Simpson index), counting effective
  species instead of raw species.
- `spaccBeta()` accumulates beta diversity and partitions it into turnover and
  nestedness components.
- `spaccCoverage()` tracks sample coverage, so curves can be standardised to a
  common completeness rather than a common site count.
- `spaccPhylo()` and `spaccFunc()` accumulate phylogenetic and functional
  diversity along the same ordering.

```{r downstream-taste, fig.width=7, fig.height=4.5, fig.alt="Hill-number accumulation curves for orders q=0, q=1, and q=2 along the spatial ordering, with the q=0 richness curve highest and the q=2 inverse-Simpson curve lowest."}
hill <- spaccHill(species, coords, q = c(0, 1, 2), n_seeds = 50, progress = FALSE, seed = 9)
plot(hill)
```

Each of these has its own vignette; the point here is that they share the
accumulation core. Choosing a `method` and `distance` configures the walk once,
and every diversity measure inherits it.

---

## Design notes

**Why several expansion methods?** No single ordering rule is correct for every
question. A growing disc (`radius`) matches a survey expanding from a point; a
chained walk (`knn`) matches a surveyor moving to the nearest accessible site; a
directional cone matches a transect along a bearing. Offering the rules as one
argument lets the analysis state its sampling model explicitly rather than
defaulting to the random null.

**Why percentile bands instead of a parametric interval?** The uncertainty in a
spatial curve is dominated by where the survey starts, which has no closed-form
distribution. Resampling the seed and reading off empirical quantiles measures
that variability directly, without assuming the curve follows a particular family.

**Why reduce abundance to presence-absence?** A species accumulation curve counts
first appearances. Whether a species has one individual or a thousand at a site
does not change whether it is new to the running total, so the accumulation core
works on presence-absence. Abundance-aware questions are answered by the Hill and
coverage extensions, which keep the counts.

**Why a hard backend switch at 500 sites?** Below a few hundred sites the
quadratic distance matrix is both small and the fastest option; above that the
memory and the per-query scan grow faster than a tree's logarithmic lookup. The
threshold is a default that `backend` overrides when a particular dataset or
benchmark calls for it.

---

## References

Arrhenius, O. (1921). Species and area. *Journal of Ecology*, 9, 95-99.

Scheiner, S. M. (2003). Six types of species-area curves. *Global Ecology and
Biogeography*, 12, 441-447.

Chiarucci, A., Bacaro, G., & Scheiner, S. M. (2011). Old and new challenges in
using species diversity for assessing biodiversity. *Philosophical Transactions of
the Royal Society B*, 366, 2426-2437.

Gotelli, N. J., & Colwell, R. K. (2001). Quantifying biodiversity: procedures and
pitfalls in the measurement and comparison of species richness. *Ecology Letters*,
4, 379-391.

Colwell, R. K., Chao, A., Gotelli, N. J., Lin, S.-Y., Mao, C. X., Chazdon, R. L.,
& Longino, J. T. (2012). Models and estimators linking individual-based and
sample-based rarefaction, extrapolation and comparison of assemblages. *Journal of
Plant Ecology*, 5, 3-21.

Chao, A., Gotelli, N. J., Hsieh, T. C., Sander, E. L., Ma, K. H., Colwell, R. K.,
& Ellison, A. M. (2014). Rarefaction and extrapolation with Hill numbers: a
framework for sampling and estimation in species diversity studies. *Ecological
Monographs*, 84, 45-67. [doi:10.1890/13-0133.1](https://doi.org/10.1890/13-0133.1)

Ugland, K. I., Gray, J. S., & Ellingsen, K. E. (2003). The species-accumulation
curve and estimation of species richness. *Journal of Animal Ecology*, 72,
888-897. [doi:10.1046/j.1365-2656.2003.00748.x](https://doi.org/10.1046/j.1365-2656.2003.00748.x)

Nekola, J. C., & White, P. S. (1999). The distance decay of similarity in
biogeography and ecology. *Journal of Biogeography*, 26, 867-878.

Shigesada, N., & Kawasaki, K. (1997). *Biological Invasions: Theory and Practice*.
Oxford University Press.

Bentley, J. L. (1975). Multidimensional binary search trees used for associative
searching. *Communications of the ACM*, 18, 509-517.

Omohundro, S. M. (1989). *Five balltree construction algorithms*. Technical Report
89-063, International Computer Science Institute, Berkeley.

Blanco, J. L., & Rai, P. K. (2014). nanoflann: a C++ header-only library for
nearest neighbor (kNN) search with k-d trees.
<https://github.com/jlblancoc/nanoflann>

## See also

- `vignette("quickstart")` -- the basic workflow end to end

- `vignette("diversity")` -- Hill numbers, beta diversity, coverage

- `vignette("extrapolation")` -- asymptotic richness models

- `vignette("spatial-analysis")` -- distance decay, endemism, zeta diversity

## Session info

```{r session-info}
sessionInfo()
```
