---
title: "Methods and Algorithms"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methods and Algorithms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 5,
  dev = "svglite",
  fig.ext = "svg",
  dev.args = list(pointsize = 14)
)
library(spacc)
set.seed(1)
```

## Overview

This vignette describes what `spacc()` computes and how. A classical species
accumulation curve draws sampling units in random order; `spacc` draws them in
spatial order, expanding outward from a focal site so the curve reflects how
richness grows with surveyed *area* rather than with effort alone. The expansion
rule is a choice, and the package offers seven of them. Each rule defines an
ordering of the sites; the curve at step $k$ is the number of distinct species
seen in the first $k$ sites of that ordering. Repeating the expansion from many
starting sites turns the spread of the curves into a confidence band.

The sections below define the vocabulary, state the accumulation problem
formally, derive each expansion method from its ordering rule, give the distance
metrics those rules depend on, map the mathematics onto the function arguments,
and explain the two-tier nearest-neighbour backend that keeps the computation
fast on large datasets. The applied vignettes build on this foundation: see
`vignette("quickstart")` for the basic workflow, `vignette("diversity")` for the
Hill, beta, and coverage extensions, and `vignette("extrapolation")` for fitting
asymptotic models to the curves.

### Contents

1. [Terminology](#terminology)

2. [Fixed-focus accumulation, concretely](#fixed-focus-accumulation-concretely)

3. [Problem formulation](#problem-formulation)

4. [Expansion methods](#expansion-methods)

5. [Distance metrics](#distance-metrics)

6. [From theory to implementation](#from-theory-to-implementation)

7. [The two-tier traversal backend](#the-two-tier-traversal-backend)

8. [Focal points and uncertainty](#focal-points-and-uncertainty)

9. [Relationship to random-order curves](#relationship-to-random-order-curves)

10. [From curves to other diversity measures](#from-curves-to-other-diversity-measures)

11. [Design notes](#design-notes)

12. [References](#references)

---

## Terminology

These terms recur throughout the package documentation.

**Site.** One sampling unit: a plot, quadrat, grid cell, or locality. Sites carry
two coordinates and a row of species records.

**Occurrence matrix.** The input `x`: an $n \times m$ matrix of $n$ sites by $m$
species. Entries are presence-absence ($0/1$) or abundance (counts). The
accumulation algorithms reduce abundance to presence-absence internally, since a
curve counts whether a species is *new*, not how many individuals it has.

**Coordinates.** The input `coords`: an $n \times 2$ table of positions, supplied
as a data frame with `x` and `y` columns, an `sf` point layer, or a precomputed
`spacc_dist` object.

**Ordering.** A permutation $\pi = (\pi_1, \dots, \pi_n)$ of the site indices that
fixes the sequence in which sites enter the curve. Every expansion method is a
rule for producing an ordering.

**Accumulation curve.** The vector $S(1), \dots, S(n)$ where $S(k)$ is the number
of distinct species recorded across the first $k$ sites of an ordering. The curve
is non-decreasing and saturates at the total species count.

**Focal point.** A continuous coordinate $c^*$ used by `knn` to rank all sites.

**Starting site.** The first site $\pi_1$ used by traversal methods.

**Neighbourhood.** The set of unvisited sites considered for the next step. For
nearest-neighbour expansion it is all unvisited sites; for the cone method it is
restricted to a directional wedge.

**Focal-point band.** The set of curves obtained from many focal points. Its
pointwise quantiles form the confidence band.

**Backend.** The data structure used to answer nearest-neighbour queries: a
precomputed distance matrix (exact) or a spatial index (k-d tree or ball tree).

---

## Fixed-focus accumulation, concretely

Start with twenty sites and one focal point. The canonical k-nearest-neighbour
ordering ranks every site by its distance from that same point.

```{r toy-points}
pts <- data.frame(
  x = c(1, 2, 1.5, 3, 3.5, 5, 5.5, 6, 2, 4,
        7, 7.5, 8, 1, 6.5, 4.5, 2.5, 8.5, 5, 3),
  y = c(1, 1.2, 2, 1, 2.5, 1, 2, 3, 3.5, 4,
        1.5, 3, 2, 4.5, 4, 3.5, 4.8, 4, 5, 5.5)
)
```

The ordering is a direct distance sort.

```{r toy-knn-order}
knn_order <- function(coords, focus) {
  order(sqrt((coords$x - focus[1])^2 + (coords$y - focus[2])^2))
}
focus <- c(3, 2.5)
ord <- knn_order(pts, focus)
```

The numbered sites expand outward from the focus.

```{r toy-knn-plot, fig.width=6, fig.height=5, fig.alt="Twenty points numbered by increasing distance from one fixed focal point."}
plot(pts$x, pts$y, pch = 19, col = "grey70", cex = 1.4,
     xlab = "x", ylab = "y", main = "Fixed-focus kNN ordering")
points(focus[1], focus[2], pch = 4, col = "#C62828", cex = 2, lwd = 2)
text(pts$x, pts$y, labels = match(seq_len(nrow(pts)), ord), pos = 3, cex = 1)
```

A different focal point gives a different ordering and curve. `spacc` samples
continuous focal points across the spatial domain to represent that variation.

---

## Problem formulation

**Input.** An occurrence matrix $X \in \{0,1,2,\dots\}^{n \times m}$ for $n$ sites
and $m$ species, and coordinates $c_1, \dots, c_n \in \mathbb{R}^2$. Let
$\mathrm{sp}(i) = \{\, j : X_{ij} > 0 \,\}$ be the set of species present at site
$i$.

**Ordering.** A method produces a permutation $\pi$ of $\{1, \dots, n\}$. For the
stochastic methods $\pi$ is a random variable whose distribution depends on the
seed and the rule.

**Accumulation function.** Given an ordering $\pi$, the curve is

$$
S_\pi(k) \;=\; \left| \, \bigcup_{t=1}^{k} \mathrm{sp}(\pi_t) \, \right|,
\qquad k = 1, \dots, n .
$$

$S_\pi$ is a step function: it rises by the count of species at $\pi_k$ that were
absent from the first $k-1$ sites, and never falls. It reaches the total richness
$S_\pi(n) = |\bigcup_i \mathrm{sp}(i)|$ regardless of the ordering, so methods
differ only in the *shape* of the approach, not the endpoint.

**Focal-point band.** Run the method from focal points $c^*_1, \dots, c^*_B$,
giving curves $S^{(1)}, \dots, S^{(B)}$
stacked as the $B \times n$ matrix returned in `$curves`. The summary at step $k$
is the across-focus mean and the empirical quantiles

$$
\bar S(k) = \frac{1}{B} \sum_{b=1}^{B} S^{(b)}(k), \qquad
\hat q_\alpha(k) = \text{quantile}_\alpha\!\left( S^{(1)}(k), \dots, S^{(B)}(k) \right).
$$

The $2.5\%$ and $97.5\%$ quantiles give the default band. The interval is not a
parametric formula; it is the sampling distribution of richness-at-effort induced
by varying the focal point.

---

## Expansion methods

The seven rules include fixed-focus expansion (`knn`), adaptive traversals
(`kncn`, `nn_walk`, `gaussian`), directional expansion (`cone`), and
geography-free baselines (`random`, `collector`).

### k-nearest neighbour (`knn`)

Sample a focal point $c^*$ in continuous space and rank all sites by distance
from that point:

$$
\pi \;=\; \operatorname{argsort}_j d(c^*, c_j).
$$

This is the fixed-focus spatially constrained rarefaction ordering described by
Chiarucci et al. (2009). By default, focal points are sampled uniformly over the
convex hull of eligible sites. An `sf` polygon supplied through `focal_domain`
represents a known irregular boundary, including holes and disconnected parts.
Exact coordinates supplied through `focal_points` reproduce a specified design.

### Nearest-neighbour walk (`nn_walk`)

From the current site $\pi_k$, move to the closest unvisited site:

$$
\pi_{k+1} \;=\; \arg\min_{j \,\notin\, \{\pi_1,\dots,\pi_k\}} d(\pi_k, j).
$$

The current site changes at every step. The traversal can follow a chain through
local clusters and cross a larger gap after a cluster is exhausted.

### k-nearest centroid neighbour (`kncn`)

Track the centroid of the visited set and move to the unvisited site closest to
it:

$$
\bar c_k = \frac{1}{k} \sum_{t=1}^{k} c_{\pi_t}, \qquad
\pi_{k+1} = \arg\min_{j \,\notin\, \{\pi_1,\dots,\pi_k\}} \| c_j - \bar c_k \| .
$$

Because the reference point is the centroid rather than the last site, the visited
set grows as a compact footprint. Its reference point updates to the centroid of
the selected sites at every step.

### Gaussian-weighted walk (`gaussian`)

A probabilistic nearest-neighbour walk. From the current site, draw the next site at random with
probability proportional to a Gaussian kernel of distance:

$$
\Pr(\pi_{k+1} = j) \;\propto\; \exp\!\left( -\frac{d(\pi_k, j)^2}{2\sigma^2} \right),
\qquad j \notin \{\pi_1,\dots,\pi_k\}.
$$

Nearby sites are favoured but not guaranteed, so the walk explores a neighbourhood
rather than always taking the single closest site. The bandwidth $\sigma$ sets how
sharp the preference is: small $\sigma$ approaches `nn_walk`, large $\sigma$ approaches
`random`. By default $\sigma$ is the median of the non-zero pairwise distances.

### Directional cone (`cone`)

Pick a random direction $\theta$ for the seed. Sites whose bearing from the seed
falls within the half-width `cone_width` of $\theta$ are accumulated first, in
order of distance; the remaining sites follow afterwards, also by distance:

$$
\text{in-cone}(j) \iff \bigl| \angle(c_j - c_s) - \theta \bigr| \le \texttt{cone\_width}.
$$

The cone models a survey that advances along a bearing, such as a transect up a
valley or along a coastline. The default half-width is $\pi/4$ (a 90-degree wedge).

### Random order (`random`)

A uniform random permutation of the sites, independent of geography. This is the
classical accumulation curve and the null model the spatial methods are measured
against.

### Collector (`collector`)

The sites in the order they appear in the data, with no randomisation. It produces
a single curve and reproduces the "collector's curve" of the order in which a
survey was actually conducted.

### Comparing the orderings

The same points, ordered by four rules, make the differences visible. Colour
encodes visiting order from first (dark) to last (light).

```{r method-schematic, fig.width=6.5, fig.height=6, fig.alt="Four panels showing the same twenty points coloured by visiting order under fixed-focus kNN, a nearest-neighbour walk, random, and collector rules."}
pal <- function(o) grDevices::hcl.colors(length(o), "Greens", rev = TRUE)[order(o)]
nn_walk_order <- function(coords, seed) {
  n <- nrow(coords); visited <- logical(n); result <- integer(n)
  current <- seed; visited[current] <- TRUE; result[1] <- current
  for (step in 2:n) {
    distances <- sqrt((coords$x - coords$x[current])^2 +
                      (coords$y - coords$y[current])^2)
    distances[visited] <- Inf
    current <- which.min(distances)
    visited[current] <- TRUE
    result[step] <- current
  }
  result
}

orders <- list(
  kNN       = knn_order(pts, focus),
  nn_walk   = nn_walk_order(pts, 1),
  random    = sample(nrow(pts)),
  collector = seq_len(nrow(pts))
)
op <- par(mfrow = c(2, 2), mar = c(2, 2, 2, 1))
for (nm in names(orders)) {
  rank_k <- match(seq_len(nrow(pts)), orders[[nm]])
  plot(pts$x, pts$y, pch = 19, cex = 1.6, col = pal(rank_k),
       xlab = "", ylab = "", main = nm, axes = FALSE); box()
}
par(op)
```

The `knn` panel expands around one point. The `nn_walk` panel follows the current
site recursively. The `random` panel has no spatial pattern, and `collector`
follows the row order.

---

## Distance metrics

The walks depend on a distance $d(i, j)$ between sites, set by the `distance`
argument.

**Euclidean.** For projected or arbitrary planar coordinates,

$$
d(i,j) = \sqrt{(x_i - x_j)^2 + (y_i - y_j)^2}.
$$

**Haversine.** For geographic coordinates given as longitude (`x`) and latitude
(`y`) in degrees, the great-circle distance in kilometres,

$$
d(i,j) = 2R \,\arcsin\!\sqrt{ \sin^2\!\tfrac{\Delta\varphi}{2}
  + \cos\varphi_i \cos\varphi_j \, \sin^2\!\tfrac{\Delta\lambda}{2} },
$$

with $\varphi$ latitude, $\lambda$ longitude, and $R = 6371$ km. Haversine needs no
projection step, so longitude-latitude data can be used directly.

**Spatiotemporal.** When sites are sampled across time, supply a `time` vector and
the distance becomes a weighted sum of a spatial and a temporal term:

$$
d(i,j) = w_{\text{space}} \, d_{\text{space}}(i,j) + w_{\text{time}} \, |t_i - t_j|.
$$

The weights `w_space` and `w_time` trade geographic against temporal proximity. A
composite distance is not a metric a spatial tree can index, so this mode always
uses the exact backend and is available for the methods that accept a distance
matrix (`nn_walk`, `gaussian`).

---

## From theory to implementation

The mathematical objects map onto arguments of `spacc()` directly.

| Concept | Argument | Effect |
|---|---|---|
| Ordering rule | `method` | Selects one of the seven expansion rules |
| Distance $d$ | `distance` | `"euclidean"` or `"haversine"` |
| Number of focal points $B$ | `n_seeds` | Curves in the focal-point band |
| Focal definition | `focal_points`, `focal_domain` | Exact foci or polygonal sampling domain for `knn` |
| Bandwidth $\sigma$ | `sigma` | Gaussian kernel width (default: median distance) |
| Cone half-width | `cone_width` | Angular wedge for `cone` (default $\pi/4$) |
| Temporal axis | `time`, `w_space`, `w_time` | Switches to composite distance |
| Fixed ordering | `order` | Supplies $\pi$ directly, bypassing `method` |
| Species split | `groups` | One curve per group, same site ordering |
| Spatial support | `support`, `include_halo` | Focal domain restricted to core sites |
| Query backend | `backend` | Nearest-neighbour engine for `nn_walk` and `kncn` |

A typical call states the rule, the metric, and the number of seeds:

```{r impl-call}
coords  <- data.frame(x = runif(120), y = runif(120))
species <- matrix(rbinom(120 * 40, 1, 0.25), nrow = 120)

sac <- spacc(species, coords, method = "knn", n_seeds = 100,
             progress = FALSE, seed = 42)
sac
```

Supplying `order` brings an externally computed ordering, for example one from a
sampling design or a different package, into the same band-and-plot machinery.
Each ordering is treated like a seed, so a matrix of orderings yields a band.

```{r impl-order}
ord_mat <- t(replicate(20, sample(nrow(species))))
sac_user <- spacc(species, coords, order = ord_mat, progress = FALSE)
sac_user$n_seeds
```

---

## The two-tier traversal backend

The `nn_walk` and `kncn` traversals repeatedly issue nearest-neighbour queries.
Two backends answer them, and `backend = "auto"` chooses by site count. Fixed-focus
`knn` computes and sorts distances from each focal point directly.

**Exact.** Precompute the full $n \times n$ distance matrix once, then answer each
query by scanning a row. The matrix costs $O(n^2)$ memory and the scan costs
$O(n)$ per step, so a full walk is $O(n^2)$. For up to a few hundred sites this is
the fastest option and the matrix fits comfortably in memory.

**Spatial tree.** Build a spatial index and query it in roughly $O(\log n)$,
avoiding the quadratic matrix entirely. For Euclidean distances `spacc` uses a
k-d tree (via nanoflann); for haversine distances it uses a ball tree, whose
spherical-cap geometry suits great-circle distance. The build is $O(n \log n)$ and
the walk is about $O(n \log n)$, which wins decisively once $n$ grows past a few
hundred.

| Backend | Distance | Memory | Per query | Selected when |
|---|---|---|---|---|
| Exact | any | $O(n^2)$ | $O(n)$ | $n \le 500$ (auto) |
| k-d tree | Euclidean | $O(n)$ | $\approx O(\log n)$ | $n > 500$ (auto) |
| Ball tree | Haversine | $O(n)$ | $\approx O(\log n)$ | $n > 500$, haversine |

The `auto` rule switches to a tree above 500 sites. The two backends compute the
same ordering rule, so they agree on the curve up to ties; only the speed differs.

```{r backend-equiv}
e <- spacc(species, coords, method = "nn_walk", backend = "exact",
           n_seeds = 30, progress = FALSE, seed = 7)
k <- spacc(species, coords, method = "nn_walk", backend = "kdtree",
           n_seeds = 30, progress = FALSE, seed = 7)
c(exact_mean_end = mean(e$curves[, ncol(e$curves)]),
  kdtree_mean_end = mean(k$curves[, ncol(k$curves)]))
```

The `gaussian` and `cone` methods use coordinates or a distance matrix directly.
The spatiotemporal composite uses the exact backend.

---

## Focal points and uncertainty

A single spatial curve is one realisation. For `knn`, `spacc` repeats the expansion
from `n_seeds` continuous focal points and reads the spread from the resulting band.
Each ordering is independent, so RcppParallel distributes them across threads and
the per-step quantiles are taken after all curves finish.

```{r seed-band, fig.width=6.5, fig.height=4.5, fig.alt="A spatial accumulation curve with a shaded confidence band. The mean curve rises and saturates; the band is widest in the early, low-effort region where the choice of starting site matters most."}
plot(sac)
```

The band is widest early and narrows as the curves converge on the shared
endpoint. Early focal locations can rank different local communities first. The
curves converge as most sites enter each ordering. More focal points improve the
Monte Carlo estimate of this spatial distribution.

```{r seed-count}
b50  <- spacc(species, coords, n_seeds = 50,  progress = FALSE, seed = 1)
b300 <- spacc(species, coords, n_seeds = 300, progress = FALSE, seed = 1)
mid <- round(ncol(b50$curves) / 4)
c(seeds_50  = diff(quantile(b50$curves[,  mid], c(.025, .975))),
  seeds_300 = diff(quantile(b300$curves[, mid], c(.025, .975))))
```

---

## Relationship to random-order curves

The classical accumulation curve in `vegan::specaccum()` draws sites in random or
collector order and averages over permutations. `spacc` reproduces that curve with
`method = "random"` or `method = "collector"` and adds the spatial methods on top,
so the two can be read against each other. The `compare()` function tests the gap
between any two curves by permutation, bootstrap, or area-under-curve.

```{r compare-random, fig.width=6.5, fig.height=4.5, fig.alt="Two accumulation curves on the same axes: the random-order curve rises faster and the spatial kNN curve lags behind it, with the shaded gap between them representing spatial turnover."}
sp <- spacc(species, coords, method = "knn",    n_seeds = 100, progress = FALSE, seed = 3)
rd <- spacc(species, coords, method = "random", n_seeds = 100, progress = FALSE, seed = 3)
cmp <- compare(sp, rd)
plot(cmp)
```

When species are aggregated, the spatial curve lags the random curve: neighbouring
sites share species, so spreading outward discovers them more slowly than mixing
distant communities at every step. The size of the gap is a measure of spatial
turnover. A spatial curve that tracks the random one signals a community with
little spatial structure; a curve that lags far behind signals strong distance
decay in composition. An existing `vegan::specaccum()` result can be imported with
`as_spacc()` to enter the same plotting and comparison machinery.

---

## From curves to other diversity measures

The expansion rule fixes an ordering. The downstream functions reuse that
ordering and change the diversity quantity accumulated at each step.

- `spaccHill()` accumulates Hill numbers of order $q = 0, 1, 2$ (richness, the
  exponential of Shannon entropy, the inverse Simpson index), counting effective
  species instead of raw species.
- `spaccBeta()` accumulates beta diversity and partitions it into turnover and
  nestedness components.
- `spaccCoverage()` tracks sample coverage, so curves can be standardised to a
  common completeness rather than a common site count.
- `spaccPhylo()` and `spaccFunc()` accumulate phylogenetic and functional
  diversity along the same ordering.

```{r downstream-taste, fig.width=6.5, fig.height=4.5, fig.alt="Hill-number accumulation curves for orders q=0, q=1, and q=2 along the spatial ordering, with the q=0 richness curve highest and the q=2 inverse-Simpson curve lowest."}
hill <- spaccHill(species, coords, q = c(0, 1, 2), n_seeds = 50, progress = FALSE, seed = 9)
plot(hill)
```

Each of these has its own vignette; the point here is that they share the
accumulation core. Choosing a `method` and `distance` configures the ordering once,
and every diversity measure inherits it.

---

## Design notes

**Why several expansion methods?** Each rule represents a sampling model.
Fixed-focus `knn` represents spatially constrained rarefaction, `kncn` represents
centroid-based compact expansion, `nn_walk` represents a nearest-neighbour
traversal, and `cone` represents directional accumulation.

**Why percentile bands instead of a parametric interval?** The uncertainty in a
spatial curve is dominated by where the survey starts, which has no closed-form
distribution. Resampling the seed and reading off empirical quantiles measures
that variability directly, without assuming the curve follows a particular family.

**Why reduce abundance to presence-absence?** A species accumulation curve counts
first appearances. Whether a species has one individual or a thousand at a site
does not change whether it is new to the running total, so the accumulation core
works on presence-absence. Abundance-aware questions are answered by the Hill and
coverage extensions, which keep the counts.

**Why a hard backend switch at 500 sites?** Below a few hundred sites the
quadratic distance matrix is both small and the fastest option; above that the
memory and the per-query scan grow faster than a tree's logarithmic lookup. The
threshold is a default that `backend` overrides when a particular dataset or
benchmark calls for it.

---

## References

Arrhenius, O. (1921). Species and area. *Journal of Ecology*, 9, 95-99.

Scheiner, S. M. (2003). Six types of species-area curves. *Global Ecology and
Biogeography*, 12, 441-447.

Chiarucci, A., Bacaro, G., Rocchini, D., Ricotta, C., Palmer, M. W., & Scheiner,
S. M. (2009). Spatially constrained rarefaction: incorporating the autocorrelated
structure of biological communities into sample-based rarefaction. *Community
Ecology*, 10, 209-214.

Gotelli, N. J., & Colwell, R. K. (2001). Quantifying biodiversity: procedures and
pitfalls in the measurement and comparison of species richness. *Ecology Letters*,
4, 379-391.

Colwell, R. K., Chao, A., Gotelli, N. J., Lin, S.-Y., Mao, C. X., Chazdon, R. L.,
& Longino, J. T. (2012). Models and estimators linking individual-based and
sample-based rarefaction, extrapolation and comparison of assemblages. *Journal of
Plant Ecology*, 5, 3-21.

Chao, A., Gotelli, N. J., Hsieh, T. C., Sander, E. L., Ma, K. H., Colwell, R. K.,
& Ellison, A. M. (2014). Rarefaction and extrapolation with Hill numbers: a
framework for sampling and estimation in species diversity studies. *Ecological
Monographs*, 84, 45-67. [doi:10.1890/13-0133.1](https://doi.org/10.1890/13-0133.1)

Ugland, K. I., Gray, J. S., & Ellingsen, K. E. (2003). The species-accumulation
curve and estimation of species richness. *Journal of Animal Ecology*, 72,
888-897. [doi:10.1046/j.1365-2656.2003.00748.x](https://doi.org/10.1046/j.1365-2656.2003.00748.x)

Nekola, J. C., & White, P. S. (1999). The distance decay of similarity in
biogeography and ecology. *Journal of Biogeography*, 26, 867-878.

Shigesada, N., & Kawasaki, K. (1997). *Biological Invasions: Theory and Practice*.
Oxford University Press.

Bentley, J. L. (1975). Multidimensional binary search trees used for associative
searching. *Communications of the ACM*, 18, 509-517.

Omohundro, S. M. (1989). *Five balltree construction algorithms*. Technical Report
89-063, International Computer Science Institute, Berkeley.

Blanco, J. L., & Rai, P. K. (2014). nanoflann: a C++ header-only library for
nearest neighbor (kNN) search with k-d trees.
<https://github.com/jlblancoc/nanoflann>

## See also

- `vignette("quickstart")` -- the basic workflow end to end

- `vignette("diversity")` -- Hill numbers, beta diversity, coverage

- `vignette("extrapolation")` -- asymptotic richness models

- `vignette("spatial-analysis")` -- distance decay, endemism, zeta diversity

## Session info

```{r session-info}
sessionInfo()
```
